The Talking Heads experiment

### 2015-04-28 chap3/chap3.tex
format relevant papers section

Similar behaviour systems and networks have
been experimented with 
for more complex tasks and it is actually the way that 
the Mars rover discussed in Chapter 1 works.\footnote{At 
the risk of simplifying, we can say that the early 
work in cybernetics (such as that of \cite{Braitenberg:1984}) 
has focused on this subsymbolic layer
and that {\itshape Artificial Intelligence}, as a research field
emerging in the late 1950's, has focused on the 
symbolic layer. Some researchers involved in a bottom up 
approach to Artificial Intelligence strongly refocused
on the subsymbolic layer. One of the most
vocal advocates of this position is \cite{Brooks:1992}.
Obviously we need the two, 
but bridging the gap is a non-trivial problem, sometimes known 
as the grounding problem. See: \cite{Harnad:1994}. }
However, the gap between the continuous dynamics of sensori-motor
intelligence and cognition remains unbridged. 
It is possible for {\itshape us} to see 
structure in the data but this structure is not perceived
nor used for control by the robot itself. 
The robot does not segment nor categorise
reality. It does not ``know'' that it is moving left or right 
and therefore cannot communicate this information to another robot. 
All processing remains at the analog continuous
level. However because it remains at this subsymbolic level, it is doubtful
whether we can ever hope to bootstrap cognitive
intelligence simply by adding more of these dynamical 
systems.\footnote{
\cite{Marr:1982} remains a classical reference outlining
the features that can be extracted, the
principled algorithms for doing it, and possible 
neurophysiological models. These are the main research
topics that still dominate research in vision. See 
\cite{Ullman:1996} for an overview.}
The difference between behavioural intelligence 
and cognitive intelligence resides in an additional 
layer of processing which is no longer continuous and analog 
but discrete and symbolic. How this second symbolic 
layer could be formed but at the same time remain 
grounded in the analog sensori-motor layer is one 
of the main research topics addressed in this book.

@@

\begin{figure}[hbtp]
\begin{center}
\includegraphics[width=0.8\columnwidth]{chap7/figs/plate-10.pdf}
\end{center}
\caption{Three examples of segmented images. The 
topic is indicated by a dashed bounding box in the 
image of the speaker. Segments which are too small 
are ignored. The topics have all been conceptualised
as being to the right and so the same word 
`gofubo' has been used to refer to them. }
\label{f:plate-10}
\end{figure}

@@

The methodology of breaking a problem into its subproblems goes quite 
a distance, but is not without danger. The processes relating
perception to language cannot be modular for reasons already mentioned. 
Each layer receives inputs which are not completely reliable and 
generates a set of possible suggestions rather than a single
``correct'' solution. Constraints have to flow
from top to bottom because there is simply not enough reliable 
information to solve the problem with a straightforward sequential 
decision process. In addition, every layer is constantly adapting 
itself to the surrounding information context. New categories 
may develop, new words need to be learned, new sensory 
channels may even emerge. So, constructing a global system is going 
to be more than simply putting together its parts. There are complex
behaviours that will only become visible when the appropriate non-modular
couplings are put in place, and this is notoriously difficult to 
do and study.

>>>>

Similar behaviour systems and networks have
been experimented with 
for more complex tasks and it is actually the way that 
the Mars rover discussed in Chapter 1 works.\footnote{At 
the risk of simplifying, we can say that the early 
work in cybernetics (such as that of \cite{Braitenberg:1984}) 
has focused on this subsymbolic layer
and that {\itshape Artificial Intelligence}, as a research field
emerging in the late 1950's, has focused on the 
symbolic layer. Some researchers involved in a bottom up 
approach to Artificial Intelligence strongly refocused
on the subsymbolic layer. One of the most
vocal advocates of this position is \cite{Brooks:1992}.
Obviously we need the two, 
but bridging the gap is a non-trivial problem, sometimes known 
as the grounding problem. See: \cite{Harnad:1994}. }
However, the gap between the continuous dynamics of sensori-motor
intelligence and cognition remains unbridged. 
It is possible for {\itshape us} to see 
structure in the data but this structure is not perceived
nor used for control by the robot itself. 
The robot does not segment nor categorise
reality. It does not \dblenquote{know} that it is moving left or right 
and therefore cannot communicate this information to another robot. 
All processing remains at the analog continuous
level. However because it remains at this subsymbolic level, it is doubtful
whether we can ever hope to bootstrap cognitive
intelligence simply by adding more of these dynamical 
systems.\footnote{
\cite{Marr:1982} remains a classical reference outlining
the features that can be extracted, the
principled algorithms for doing it, and possible 
neurophysiological models. These are the main research
topics that still dominate research in vision. See 
\cite{Ullman:1996} for an overview.}
The difference between behavioural intelligence 
and cognitive intelligence resides in an additional 
layer of processing which is no longer continuous and analog 
but discrete and symbolic. How this second symbolic 
layer could be formed but at the same time remain 
grounded in the analog sensori-motor layer is one 
of the main research topics addressed in this book.

@@

\begin{figure}[hbtp]
\begin{center}
\includegraphics[width=0.8\columnwidth]{chap7/figs/plate-10.pdf}
\end{center}
\caption{Three examples of segmented images. The 
topic is indicated by a dashed bounding box in the 
image of the speaker. Segments which are too small 
are ignored. The topics have all been conceptualised
as being to the right and so the same word 
\smplenquote{gofubo} has been used to refer to them. }
\label{f:plate-10}
\end{figure}

@@

The methodology of breaking a problem into its subproblems goes quite 
a distance, but is not without danger. The processes relating
perception to language cannot be modular for reasons already mentioned. 
Each layer receives inputs which are not completely reliable and 
generates a set of possible suggestions rather than a single
\dblenquote{correct} solution. Constraints have to flow
from top to bottom because there is simply not enough reliable 
information to solve the problem with a straightforward sequential 
decision process. In addition, every layer is constantly adapting 
itself to the surrounding information context. New categories 
may develop, new words need to be learned, new sensory 
channels may even emerge. So, constructing a global system is going 
to be more than simply putting together its parts. There are complex
behaviours that will only become visible when the appropriate non-modular
couplings are put in place, and this is notoriously difficult to 
do and study.

### 2015-04-28 chap4/chap4.tex
format relevant papers section

The previous chapter laid the perceptual groundwork 
for investigating conceptualisation and language 
communication. It proposed a set of mechanisms for segmenting a 
scene and extracting characteristics about each 
segment. The next task of the agent is to use these results
from the perceptual layer to categorise 
and hence conceptualise the
scene. For the time being, I will 
ignore the full complexity of natural
language meaning and focus only on the most 
simple type of conceptualisation one can
imagine, namely categories, which logicians refer 
to as unary predicates, as well as conjunctive 
combinations of categories. Words like `blue', `light' or
`square' name such categories. We not only need a way 
that the Talking Heads can use such categories to conceptualise
a scene based on visual perception, but we also need 
mechanisms that can explain how such categories might 
develop or be acquired by each agent autonomously and 
without supervised training. This is clearly an enormous
challenge and no known universally accepted solution 
exists to this problem.

@@

Then there is a third paradox, first suggested
by Jean-Jacques Rousseau. In order to communicate,
the speaker and the hearer need to share the building
blocks of the conceptualisations underlying their communication. 
If I say ``the wine glass on the table'', I expect 
the hearer to be able to recognise what objects are 
tables, when something is on the table or not on the table, and 
when something is a wine glass versus another kind 
of glass. But if every individual learns categories independently
and autonomously, how can they ever become shared? It is 
assumed that language helps in establishing a shared ontology within 
a language community but language itself depends on 
shared categories, so how can the whole system ever 
get off the ground; how can this chicken and egg situation
be broken?

@@

\is{innate categories}This innateness hypothesis suggests that the brain comes
with {\it categorisation organs}, small neuronal circuits
capable of performing 
the mapping of some idealised universal form onto
reality. Consequently the human genome
must include a set of {\it concept genes} which regulate how each of 
these categorisation organs should grow during 
development. Rationalists claim
that it is absurd to think of categories as being 
learned from example cases supplied by the environment, just 
as it is absurd to say that the hand learns to 
grow five fingers.

@@

There is something to say both for a rationalist and 
for an empiricist approach, indeed otherwise so many
serious thinkers could not have believed fervently 
in one position or the other. Rationalists point to the 
fact that children acquire concepts
surprisingly quickly and apparently with very little stimuli
and that anthropological observations have 
shown that there are strong universal tendencies
for basic perceptually grounded categories, such as colour or 
space. However, more detailed observations show
that the acquisition of categories in children goes 
in fact very slowly. For example, concepts like cause-effect
and the correct use of the word `because', the proper
use of tenses, etc. all take years to develop. 
Adults keep acquiring
new categories through out their lifetime, which makes 
it difficult to maintain that they are part of the 
human genome. For example, 
airline pilots and sailors categorise the direction and 
strength of winds and the shapes and colours of clouds, 
so as to predict turbulences, future weather, or 
advantageous trajectories. There are
occasionally profound differences in how cultures conceptualise
reality. These differences do not seem to be innate because
everybody who has had sufficient exposure to the 
culture, preferably at an early age and therefore without too much 
preconception, can normally acquire them.\footnote{
Rationalists often argue that there is no other way 
to explain the child's rapid acquisition of concepts
but detailed psycholinguistic observations have shown that 
child language understanding (which is the 
clearest sign whether certain concepts have been acquired)
is often deceptive because 
non-verbal strategies may lead to appropriate answers
to adult questions and thus the appearance of 
understanding.}

@@

The difference between an animal
reacting to a limited set of environmental stimuli with 
a rigid neural apparatus and a cognitively endowed
human being is precisely the high degree of flexibility and 
adaptivity of the latter.
It is therefore not surprising that clear-cut 
{\it categorisation organs} which have the same structure in 
all humans and are located at the same position in the 
brain have not been found. The microstructure of the
brain does not consist of
neatly separated organs and it therefore does not make 
sense to look for genes that regulate their maturation. 
Neurobiologists tell us that 
the brain appears more like an organically grown 
tissue rather than a delicately tuned machine laid
out by genetic programs. In contrast to insect brains or 
brains of lower animals, the mammalian brain is 
capable of regenerating itself to some extent after
damage, and brain tissue
from one higher order animal can be implanted in another
one, causing a resurgence of lost function, even if the 
source location of the transplant is different. 
For example, if tissue from the visual cortex is
implanted in the auditory cortex, it will 
regenerate and function as part of the auditory 
cortex. Pathways to the visual cortex can be
redirected to the auditory cortex, which will cause 
the auditory cortex to take on functions of visual processing.

@@

Note that selectionism is {\it not} applied at 
the level of a species as in biological evolution but 
at the level of the cognitive structures in each individual
as he or she develops and adapts during his or her life 
time. The parallel exploration and the competition of 
alternative ways to categorise reality take 
place in a single individual interacting with the environment
and therefore are very rapid.

@@

In a selectionist approach, a repertoire of primitive operations
is given, presumably implemented by the 
basic biochemistry of the neural systems, and there are
ways to combine these operations into more complex
{\it visual programs}. Those
programs yielding an outcome which
is afterwards used successfully in categorisation are 
retained and the others discarded, thus establishing 
a co-evolution between a repertoire of sensory channels
and a repertoire of 
categories. We have already done some successful experiments 
in this direction, see \cite{Dejong:1999},  
but this theme will not be pursued further here as a full discussion 
would be too much a digression from the main line
of investigation. For the remainder of this book,
the sensory channels will be pre-programmed, 
although it is still up to the selectionist 
categorisation process to discover which ones are useful
in the environments presented to the agents and which 
ones are not.

@@

Consider the category [\textsc{large}]. What do all large 
objects have in common? At first sight very little. 
Almost any physical object can be called `large' in 
one context or another. 
It is going to be very difficult for a learning agent
to determine some commonality, even if given thousands 
of examples of large objects. In the beginning, the agent
might be confused by commonalities in 
colour or position or shape. 
Only with some luck, will the learning algorithm
start to zoom in on size. 
This is what makes inductive learning so slow and 
why it is criticised, rightfully, by rationalists. 
In fact, searching for commonality hardly makes sense for
many categories. [\textsc{large}] is only meaningful in
opposition to [\textsc{small}], [\textsc{left}] only makes sense
in opposition to [\textsc{right}]. Often categorisation
is relative to the context (a block which is large when surrounded
by smaller blocks may be categorised as
small when surrounded by larger
blocks) and occasionally it is relative to the objects themselves
(a large mouse is always categorised as
much smaller than a small elephant).

@@

A realistic agent has hundreds of sensory
channels, and humans probably have tens of thousands
of them, probably grouped with respect to the domains
to which they apply.\footnote{
This is strongly suggested by psychological 
data on the presence of conceptual spaces in human 
categorisation. See \cite{Gardenfors:1999}.} There are discrimination trees
for each of 
these channels or for combinations of them, and 
each can possibly yield a distinctive
category. The categorisation process can therefore be
envisioned like a huge Pachinko hall, in which
balls are trickling down in parallel in hundreds or 
thousands of machines. It is highly 
likely that more than one solution is found
when there are a lot of trees, particularly 
if combinations of categories are allowed as well. 
So, an additional competitive process must 
take place to rank categories even 
though multiple solutions are offered to subsequent
verbalisation processes. We see therefore the same
characteristics as for the perceptual layer, and thus
the ``sieve architecture'' also 
applies for the conceptual layer (\figref{f:sieve}).

@@

Still another way to categorise reality is by imposing 
an order on the segments based on their values for 
a particular sensory channel. For example, we can 
order the segments based on the \textsc{hpos} channel (i.e. 
from left to right) and then introduce relational categories
like `left of one segment in the series', or `the left-most object'. 
Similarly we can order the segments based on the 
\textsc{height} channel (i.e. in terms of their size) and then 
have categories that select the smallest (i.e. the 
first segment in this ordering), or those greater than 
some other one.

>>>>

The previous chapter laid the perceptual groundwork 
for investigating conceptualisation and language 
communication. It proposed a set of mechanisms for segmenting a 
scene and extracting characteristics about each 
segment. The next task of the agent is to use these results
from the perceptual layer to categorise 
and hence conceptualise the
scene. For the time being, I will 
ignore the full complexity of natural
language meaning and focus only on the most 
simple type of conceptualisation one can
imagine, namely categories, which logicians refer 
to as unary predicates, as well as conjunctive 
combinations of categories. Words like \smplenquote{blue}, \smplenquote{light} or
\smplenquote{square} name such categories. We not only need a way 
that the Talking Heads can use such categories to conceptualise
a scene based on visual perception, but we also need 
mechanisms that can explain how such categories might 
develop or be acquired by each agent autonomously and 
without supervised training. This is clearly an enormous
challenge and no known universally accepted solution 
exists to this problem.

@@

Then there is a third paradox, first suggested
by Jean-Jacques Rousseau. In order to communicate,
the speaker and the hearer need to share the building
blocks of the conceptualisations underlying their communication. 
If I say \dblenquote{the wine glass on the table}, I expect 
the hearer to be able to recognise what objects are 
tables, when something is on the table or not on the table, and 
when something is a wine glass versus another kind 
of glass. But if every individual learns categories independently
and autonomously, how can they ever become shared? It is 
assumed that language helps in establishing a shared ontology within 
a language community but language itself depends on 
shared categories, so how can the whole system ever 
get off the ground; how can this chicken and egg situation
be broken?

@@

\is{innate categories}This innateness hypothesis suggests that the brain comes
with {\itshape categorisation organs}, small neuronal circuits
capable of performing 
the mapping of some idealised universal form onto
reality. Consequently the human genome
must include a set of {\itshape concept genes} which regulate how each of 
these categorisation organs should grow during 
development. Rationalists claim
that it is absurd to think of categories as being 
learned from example cases supplied by the environment, just 
as it is absurd to say that the hand learns to 
grow five fingers.

@@

There is something to say both for a rationalist and 
for an empiricist approach, indeed otherwise so many
serious thinkers could not have believed fervently 
in one position or the other. Rationalists point to the 
fact that children acquire concepts
surprisingly quickly and apparently with very little stimuli
and that anthropological observations have 
shown that there are strong universal tendencies
for basic perceptually grounded categories, such as colour or 
space. However, more detailed observations show
that the acquisition of categories in children goes 
in fact very slowly. For example, concepts like cause-effect
and the correct use of the word \smplenquote{because}, the proper
use of tenses, etc. all take years to develop. 
Adults keep acquiring
new categories through out their lifetime, which makes 
it difficult to maintain that they are part of the 
human genome. For example, 
airline pilots and sailors categorise the direction and 
strength of winds and the shapes and colours of clouds, 
so as to predict turbulences, future weather, or 
advantageous trajectories. There are
occasionally profound differences in how cultures conceptualise
reality. These differences do not seem to be innate because
everybody who has had sufficient exposure to the 
culture, preferably at an early age and therefore without too much 
preconception, can normally acquire them.\footnote{
Rationalists often argue that there is no other way 
to explain the child's rapid acquisition of concepts
but detailed psycholinguistic observations have shown that 
child language understanding (which is the 
clearest sign whether certain concepts have been acquired)
is often deceptive because 
non-verbal strategies may lead to appropriate answers
to adult questions and thus the appearance of 
understanding.}

@@

The difference between an animal
reacting to a limited set of environmental stimuli with 
a rigid neural apparatus and a cognitively endowed
human being is precisely the high degree of flexibility and 
adaptivity of the latter.
It is therefore not surprising that clear-cut 
{\itshape categorisation organs} which have the same structure in 
all humans and are located at the same position in the 
brain have not been found. The microstructure of the
brain does not consist of
neatly separated organs and it therefore does not make 
sense to look for genes that regulate their maturation. 
Neurobiologists tell us that 
the brain appears more like an organically grown 
tissue rather than a delicately tuned machine laid
out by genetic programs. In contrast to insect brains or 
brains of lower animals, the mammalian brain is 
capable of regenerating itself to some extent after
damage, and brain tissue
from one higher order animal can be implanted in another
one, causing a resurgence of lost function, even if the 
source location of the transplant is different. 
For example, if tissue from the visual cortex is
implanted in the auditory cortex, it will 
regenerate and function as part of the auditory 
cortex. Pathways to the visual cortex can be
redirected to the auditory cortex, which will cause 
the auditory cortex to take on functions of visual processing.

@@

Note that selectionism is {\itshape not} applied at 
the level of a species as in biological evolution but 
at the level of the cognitive structures in each individual
as he or she develops and adapts during his or her life 
time. The parallel exploration and the competition of 
alternative ways to categorise reality take 
place in a single individual interacting with the environment
and therefore are very rapid.

@@

In a selectionist approach, a repertoire of primitive operations
is given, presumably implemented by the 
basic biochemistry of the neural systems, and there are
ways to combine these operations into more complex
{\itshape visual programs}. Those
programs yielding an outcome which
is afterwards used successfully in categorisation are 
retained and the others discarded, thus establishing 
a co-evolution between a repertoire of sensory channels
and a repertoire of 
categories. We have already done some successful experiments 
in this direction, see \cite{Dejong:1999},  
but this theme will not be pursued further here as a full discussion 
would be too much a digression from the main line
of investigation. For the remainder of this book,
the sensory channels will be pre-programmed, 
although it is still up to the selectionist 
categorisation process to discover which ones are useful
in the environments presented to the agents and which 
ones are not.

@@

Consider the category [\textsc{large}]. What do all large 
objects have in common? At first sight very little. 
Almost any physical object can be called \smplenquote{large} in 
one context or another. 
It is going to be very difficult for a learning agent
to determine some commonality, even if given thousands 
of examples of large objects. In the beginning, the agent
might be confused by commonalities in 
colour or position or shape. 
Only with some luck, will the learning algorithm
start to zoom in on size. 
This is what makes inductive learning so slow and 
why it is criticised, rightfully, by rationalists. 
In fact, searching for commonality hardly makes sense for
many categories. [\textsc{large}] is only meaningful in
opposition to [\textsc{small}], [\textsc{left}] only makes sense
in opposition to [\textsc{right}]. Often categorisation
is relative to the context (a block which is large when surrounded
by smaller blocks may be categorised as
small when surrounded by larger
blocks) and occasionally it is relative to the objects themselves
(a large mouse is always categorised as
much smaller than a small elephant).

@@

A realistic agent has hundreds of sensory
channels, and humans probably have tens of thousands
of them, probably grouped with respect to the domains
to which they apply.\footnote{
This is strongly suggested by psychological 
data on the presence of conceptual spaces in human 
categorisation. See \cite{Gardenfors:1999}.} There are discrimination trees
for each of 
these channels or for combinations of them, and 
each can possibly yield a distinctive
category. The categorisation process can therefore be
envisioned like a huge Pachinko hall, in which
balls are trickling down in parallel in hundreds or 
thousands of machines. It is highly 
likely that more than one solution is found
when there are a lot of trees, particularly 
if combinations of categories are allowed as well. 
So, an additional competitive process must 
take place to rank categories even 
though multiple solutions are offered to subsequent
verbalisation processes. We see therefore the same
characteristics as for the perceptual layer, and thus
the \dblenquote{sieve architecture} also 
applies for the conceptual layer (\figref{f:sieve}).

@@

Still another way to categorise reality is by imposing 
an order on the segments based on their values for 
a particular sensory channel. For example, we can 
order the segments based on the \textsc{hpos} channel (i.e. 
from left to right) and then introduce relational categories
like \smplenquote{left of one segment in the series}, or \smplenquote{the left-most object}. 
Similarly we can order the segments based on the 
\textsc{height} channel (i.e. in terms of their size) and then 
have categories that select the smallest (i.e. the 
first segment in this ordering), or those greater than 
some other one.
